# Computing

```{r, include = FALSE} 
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE, fig.align = "center")
```

This is not a book about R. It is however, a book that uses R. Because of this, you will need to be familiar with R. The text will point out some thing about R along the way, but some previous study of R is necessary.

The following (freely availible) readings are highly recommended:

- [Hands-On Programming with R](https://rstudio-education.github.io/hopr/) - *Garrett Grolemund*
    - If you have never used R or RStudio before, Part 1, Chapters 1 - 3, will be useful.
- [R for Data Science](https://r4ds.had.co.nz/) - *Garrett Grolemund, Hadley Wickham*
    - This book helps getting you up to speed working with data in R. While it is a lot of reading, Chapters 1 - 21 are highly recommended. 
- [Advanced R](https://adv-r.hadley.nz/) - *Hadley Wickham*
    - Part I, Chapters 1 - 8, of this book will help create a mental model for working with R. These chapters are not an easy read, so they should be returned to often. (Chapter 2 could be safely skipped for our purposes, but is important if you will use R in the long term.)

If you are a UIUC student who took the course STAT 420, the first six chapters of that book could serve as a nice refresher.

- [Applied Statistics with R](https://daviddalpiaz.github.io/appliedstats/) - *David Dalpiaz*

***

## Resources

The following resources are more specific or more advanced, but could still prove to be useful.

### R

- [Efficient R programming](https://csgillespie.github.io/efficientR/)
- [R Programming for Data Science](https://bookdown.org/rdpeng/rprogdatascience/)
- [R Graphics Cookbook](https://r-graphics.org/)
- [Modern Dive](https://moderndive.com/index.html)
- [The `tidyverse` Website](tidyverse.org/)
    - [`dplyr` Website](https://dplyr.tidyverse.org/)
    - [`readr` Website](https://readr.tidyverse.org/)
    - [`tibble` Website](https://tibble.tidyverse.org/)
    - [`forcats` Website](https://forcats.tidyverse.org/)

### RStudio

- [RStudio IDE Cheatsheet](https://resources.rstudio.com/rstudio-cheatsheets/rstudio-ide-cheat-sheet)
- [RStudio Resources](https://resources.rstudio.com/)

### R Markdown

- [R Markdown Cheatsheet](https://resources.rstudio.com/rstudio-cheatsheets/rmarkdown-2-0-cheat-sheet)
- [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/) - *Yihui Xie, J. J. Allaire, Garrett Grolemund*
- [R4DS R Markdown Chapter](https://r4ds.had.co.nz/r-markdown.html)

#### Markdown

- [Daring Fireball - Markdown: Basics](https://daringfireball.net/projects/markdown/basics)
- [GitHub - Mastering Markdown](https://guides.github.com/features/mastering-markdown/)
- [CommonMark](https://commonmark.org/)

***

## BSL Idioms

Things here supercede everythign above.

### Reference Style

- [`tidyverse` Style Guide](https://style.tidyverse.org/)

### BSL Style Overrides

- TODO: `=` instead of `<-`
    - http://thecoatlessprofessor.com/programming/an-opinionated-tale-of-why-you-should-replace---with-/
- TODO: never use `T` or `F`, only `TRUE` or `FALSE`

```{r}
FALSE == TRUE
F     == TRUE
F     =  TRUE
F     == TRUE
```

- TODO: never ever ever use `attach()`
- TODO: never ever ever use `<<-`
- TODO: never ever ever use `setwd()` or set a working directory some other way
- TODO: a newline before and after any chunk
- TODO: use headers appropriately! (short names, good structure)
- TODO: never ever ever put spaces in filenames. use `-`. (others will use `_`.)
- TODO: load all needed packages at the beginning of an analysis in a single chunk (TODO: pros and cons of this approach)
- TODO: one plot per chunk! no other printed output

Be consistent...

- with yourself!
- with your group!
- with your organization!

```{r, eval = FALSE}
set.seed(1337);mu=10;sample_size=50;samples=100000;
x_bars=rep(0, samples)
for(i in 1:samples)
{
x_bars[i]=mean(rpois(sample_size,lambda = mu))}
x_bar_hist=hist(x_bars,breaks=50,main="Histogram of Sample Means",xlab="Sample Means",col="darkorange",border = "dodgerblue")
mean(x_bars>mu-2*sqrt(mu)/sqrt(sample_size)&x_bars<mu+2*sqrt(mu)/sqrt(sample_size))
```

### Objects and Functions

> To understand computations in R, two slogans are helpful:
> 
> - Everything that exists is an object. 
> - Everything tha thappens is a function call. 
>
> — John Chambers

### Print versus Return

```{r}
cars_mod = lm(dist ~ speed, data = cars)
```

```{r}
summary(cars_mod)
```

```{r}
is.list(summary(cars_mod))
```

```{r}
names(summary(cars_mod))
```

```{r}
str(summary(cars_mod))
```

```{r, eval = FALSE}
# RStudio only
View(summary(cars_mod))
```

### Help

- TODO: `?`, google, stack overflow, (office hours, course forums)

### Keyboard Shortcuts

- TODO: copy-paste, switch program, switch tab, etc...
- TODO: TAB!!!
- TODO: new chunk!
- TODO: style!
- TODO: keyboard shortcut for keyboard shortcut

***
