# Classification

## STAT 432 Materials

- [**Slides** | Classification: Introduction](https://fall-2019.stat432.org/slides/classification.pdf)
- [**Code** | Some Classification Code](https://fall-2019.stat432.org/misc/some-class-code-for-class.R)
- [**Slides** | Classification: Binary Classification](https://fall-2019.stat432.org/slides/binary-classification.pdf)
- [**Code** | Some Binary Classification Code](https://fall-2019.stat432.org/misc/some-binary-class-code-for-class.R)
- [**Slides** | Classification: Nonparametric Classification](https://fall-2019.stat432.org/slides/nonparametric-classification.pdf)
- [**Reading** | STAT 420: Logistic Regression](https://daviddalpiaz.github.io/appliedstats/logistic-regression.html)

***

## Bayes Classifier

- TODO: Not the same as naïve Bayes classifier

$$
C^B(x) = \underset{k \in \{1, 2, \ldots K\}}{\text{argmax}} P\left[ Y = k \mid X = x \right]
$$

***

### Bayes Error Rate

$$
1 - \mathbb{E}\left[ \underset{k}{\text{max}} \ P[Y = k \mid X] \right]
$$

- TODO: https://topepo.github.io/caret/visualizations.html
- TODO: https://en.wikipedia.org/wiki/Confusion_matrix
- TODO: https://en.wikipedia.org/wiki/Matthews_correlation_coefficient
- TODO: https://people.inf.elte.hu/kiss/11dwhdm/roc.pdf
- TODO: https://www.cs.cmu.edu/~tom/mlbook/NBayesLogReg.pdf
- TODO: http://www.oranlooney.com/post/viz-tsne/
- TODO: https://web.expasy.org/pROC/
- TODO: https://bmcbioinformatics.biomedcentral.com/track/pdf/10.1186/1471-2105-12-77
- TODO: https://en.wikipedia.org/wiki/Receiver_operating_characteristic
- TODO: https://papers.nips.cc/paper/2020-on-discriminative-vs-generative-classifiers-a-comparison-of-logistic-regression-and-naive-bayes.pdf

```{r}
library(tibble)
library(caret)
library(rpart)
library(nnet)
```

```{r}
gen_data = function(n = 100) {
  x = sample(c(0, 1), prob = c(0.4, 0.6), size = n, replace = TRUE)
  y = ifelse(test = {x == 0},
             yes = sample(c("A", "B", "C"), size = n, prob = c(0.25, 0.50, 0.25), replace = TRUE),
             no = sample(c("A", "B", "C"), size = n, prob = c(0.1, 0.1, 0.4) / 0.6, replace = TRUE))

  tibble(x = x, y = factor(y))
}

test_cases = tibble(x = c(0, 1))

set.seed(42)
some_data = gen_data()

predict(knn3(y ~ x, data = some_data), test_cases)
predict(rpart(y ~ x, data = some_data), test_cases)
predict(nnet(y ~ x, data = some_data, size = 0, skip = TRUE, trace = FALSE), test_cases)
```

## Modeling

### Linear Models

### k-Nearest Neighbors

### Decision Trees
