# Classification: Handwriting

```{r, include = FALSE} 
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE, fig.align = "center")
```

```{r, message = FALSE, warning = FALSE}
library(readr)
library(tibble)
library(dplyr)
library(purrr)
library(ggplot2)
library(ggridges)
library(lubridate)
library(randomForest)
library(rpart)
library(rpart.plot)
library(cluster)
library(caret)
library(factoextra)
library(rsample)
library(janitor)
library(rvest)
library(dendextend)
library(knitr)
library(kableExtra)
library(ggthemes)
```

```{r, echo = FALSE}
theme_set(new = theme_light())
```

- TODO: Show package messaging? check conflicts!
- TODO: Should this be split into three analyses with different packages?

## Background

- TODO: https://en.wikipedia.org/wiki/MNIST_database
- TODO: http://yann.lecun.com/exdb/mnist/

## Data

- TODO: How is this data pre-processed?
- TODO: https://gist.github.com/daviddalpiaz/ae62ae5ccd0bada4b9acd6dbc9008706
- TODO: https://github.com/itsrainingdata/mnistR
- TODO: https://pjreddie.com/projects/mnist-in-csv/
- TODO: http://varianceexplained.org/r/digit-eda/

```{r, load-mnist-remote, warning = FALSE, message = FALSE, eval = FALSE, echo = FALSE}
mnist_trn = read_csv(
  file = "https://pjreddie.com/media/files/mnist_train.csv", 
  col_names = FALSE)

mnist_tst = read_csv(
  file = "https://pjreddie.com/media/files/mnist_test.csv", 
  col_names = FALSE)
```

```{r, eval = FALSE, echo = FALSE}
write_csv(mnist_trn[1:1000, ], path = "data/mnist_train_subest.csv")
write_csv(mnist_tst, path = "data/mnist_test.csv")
```

```{r, clean-mnist, echo = FALSE, eval = FALSE}
rm(mnist_trn)
rm(mnist_tst)
```

```{r, load-mnist-local, warning = FALSE, message = FALSE,}
mnist_trn = read_csv(file = "data/mnist_train_subest.csv")
mnist_tst = read_csv(file = "data/mnist_test.csv")
```

```{r, split-mnist}
mnist_trn_y = as.factor(mnist_trn$X1)
mnist_tst_y = as.factor(mnist_tst$X1)

mnist_trn_x = mnist_trn[, -1]
mnist_tst_x = mnist_tst[, -1]
```

```{r, clean-mnist-local, echo = FALSE}
rm(mnist_trn)
rm(mnist_tst)
```

- TODO: If we were going to tune a model, we would need a validation split as well. We're going to be lazy and just fit a single random forest.
- TODO: This is an agreed upon split.

## EDA

```{r}
pixel_positions = expand.grid(j = sprintf("%02.0f", 1:28), 
                              i = sprintf("%02.0f", 1:28))
pixel_names = paste("pixel", pixel_positions$i, pixel_positions$j, sep = "-")
```

```{r}
colnames(mnist_trn_x) = pixel_names
colnames(mnist_tst_x) = pixel_names
```

```{r}
show_digit = function(arr784, col = gray(12:1 / 12), ...) {
  image(matrix(as.matrix(arr784), nrow = 28)[, 28:1], 
        col = col, xaxt = "n", yaxt = "n", ...)
  grid(nx = 28, ny = 28)
}
```

```{r, plot-digits, echo = FALSE, fig.height = 6, fig.width = 14}
par(mfrow = c(2, 5))
show_digit(mnist_trn_x[2, ],  main = "An Example Zero")
show_digit(mnist_trn_x[4, ],  main = "An Example One")
show_digit(mnist_trn_x[6, ],  main = "An Example Two")
show_digit(mnist_trn_x[8, ],  main = "An Example Three")
show_digit(mnist_trn_x[10, ], main = "An Example Four")
show_digit(mnist_trn_x[1, ],  main = "An Example Five")
show_digit(mnist_trn_x[14, ], main = "An Example Six")
show_digit(mnist_trn_x[16, ], main = "An Example Seven")
show_digit(mnist_trn_x[18, ], main = "An Example Eight")
show_digit(mnist_trn_x[5, ],  main = "An Example Nine")
```

## Modeling

```{r}
set.seed(42)
mnist_rf = randomForest(x = mnist_trn_x, y = mnist_trn_y, ntree = 100)
```

## Model Evaluation

```{r}
mnist_tst_pred = predict(mnist_rf, mnist_tst_x)
mean(mnist_tst_pred == mnist_tst_y)
```

```{r}
table(predicted = mnist_tst_pred, actual = mnist_tst_y)
```

## Discussion

```{r, echo = FALSE}
plot_mistake = function(actual, predicted) {
  
  obs_to_plot = which(mnist_tst_y == actual & mnist_tst_pred == predicted)[1:9]
  show_digit(mnist_tst_x[obs_to_plot[1], ])
  show_digit(mnist_tst_x[obs_to_plot[2], ])
  show_digit(mnist_tst_x[obs_to_plot[3], ])
  show_digit(mnist_tst_x[obs_to_plot[4], ])
  show_digit(mnist_tst_x[obs_to_plot[5], ])
  show_digit(mnist_tst_x[obs_to_plot[6], ])
  show_digit(mnist_tst_x[obs_to_plot[7], ])
  show_digit(mnist_tst_x[obs_to_plot[8], ])
  show_digit(mnist_tst_x[obs_to_plot[9], ])
}
```

```{r, fig.height = 12, fig.width = 12}
par(mfrow = c(3, 3))
plot_mistake(actual = 6, predicted = 4)
```

```{r}
mnist_obs_to_check = 2
predict(mnist_rf, mnist_tst_x[mnist_obs_to_check, ], type = "prob")[1, ]
mnist_tst_y[mnist_obs_to_check]
```

```{r, fig.height = 4, fig.width = 4}
show_digit(mnist_tst_x[mnist_obs_to_check, ])
```
